A multi-scale story of the diffusion of a new technology: the web

### 2025-03-12 paper/submission2/anonymised.tex
paper v2

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A multi-scale story of the diffusion of a new technology: the web},
  pdfauthor={Emmanouil Tranos},
  pdfkeywords={diffusion, technologies, Web, digital},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
Despite the broad consensus that differences in technological adoption
and sophistication can lead to uneven trajectories of local development
as new technologies can increase productivity and allow firms to
introduce new products and services, economic geographers have shied
away from exploring the diffusion of digital technologies across space
and time. This paper fills this gap by exposing the roles of different
spatial mechanisms in the diffusion of an intangible, digital technology
at very granular geographical scales. It does so by employing a novel
source of data and machine learning algorithms to model the active
engagement with Web during 1996-2012.
\end{abstract}

@@

Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of
\citet{hagerstrand1968innovation}, which demonstrated the importance of
communications as a mechanism for the diffusion of innovations, being
the most impactful. It was followed by various empirical studies -- see
for instance \citet{ormrod1990} and \citet{iso2005}. However, the torch
of exploring and modelling such diffusion processes had been passed to
other disciplines such as economics, business studies and sociology well
before the `cultural turn' of economic geography
\citep{perkins2005international, ding2010modeling}. As a consequence,
relevant studies did not investigate diffusion mechanisms at granular
geographical scales to provide local policy insights, for instance,
regarding infrastructural d evelopment and local technology policies. A
potential explanation of the lack of geographical studies exploring the
diffusion of new and, more specifically for this paper, digital
technologies across both space and time can be attributed to the
scarcity of relevant and granular enough data, a problem also
highlighted by \citet{iso2005} and \citet{kemeny2011international}. As
\citet{zook2022mapping} pointed, digital activities are hardly ever
captured in official data.

This paper offers such a contribution: an economic geography study
revealing the importance of spatial mechanisms for the diffusion of an
intangible, digital technology that is the Web in the UK at a high level
of spatial granularity from the very early days until its maturity
(1996-2012). It specifically focuses on the diffusion of commercial
websites. It does so by employing a novel source of big data and machine
learning algorithms to model the \emph{active} engagement with Web --
that is creating and maintaining instead of just browsing a website --
during that period and tests how different spatial and temporal
mechanisms shaped its diffusion. This paper also exemplifies how the
combination of data sources and state-of-the-art research methods which
escape the traditional social science domain can offer new lenses to
geographical research regarding the understanding of technological
diffusion.

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffuse over space and time and use this knowledge to make predictions
about the diffusion of related future technologies. As per
\citet{leibowicz2016representing}, historical studies agree that
technologies diffuse differently in terms of times, rates, and
geographies and can be driven by related policies \citep{victor1993}.
\citet{meade2021modelling} highlight that a variety of actors have a
direct interest in gaining such knowledge including network equipment
suppliers, network operators, regulatory and local authorities. These
processes and their effects vary a lot across scales: although the
diffusion of a new technology might not be optimal at a local level, it
might be beneficial from a global perspective as it could lead to faster
diffusion to less advantaged places \citep{leibowicz2016representing}.
Despite the spatial heterogeneity of such diffusion mechanisms and the
policy relevance, there is a scarcity of studies analysing the diffusion
of new digital technologies at a detailed geographical level because of
the lack of detailed enough data and adequate research methods.

@@

Another line of motivation has to do with the very nature of the Web as
a digital technology and how it evolved over time. The digital
geographies field has debunked premature claims about the Internet's
placeless-ness by mapping various of its facets and technological layers
\citep{tranos2013death}, but the lack of relevant long time-series data
prevented researchers from observing digital phenomena overtime. Most of
these studies focused on its hard infrastructure
\citep[e.g.][]{malecki2002economic, moss2000internet}, on Internet
subscribers \citep[e.g.][]{blank2018local}, or on social media users
\citep[e.g.][]{crampton2013beyond}. Exceptions include early work from
\citet{zook2000web}, \citet{zook2000web} and \citet{moss1997tracking},
which utilised domain name registration data and use it to map and
illustrate the spatial footprint of the web content. Pivotal as these
contributions have been in understanding the economic geographies of the
Internet, they were based on cross-sectional, often spatially detailed,
mapping and on shallow time comparisons (e.g.~1993-1998) and did not
observe the evolution and diffusion of the Web until its maturity. As we
know from the relevant literature, different technologies diffuse at
different rates and there is no theory that can explain these
differences \citep{leibowicz2016representing}. Hence, mapping and
modelling the diffusion of the Web among firms can offer useful
insights.

@@

To analyse these data, a diverse set of methods are employed. Firstly,
cumulative adoption curves are estimated for the UK as a whole and for
the Local Authority Districts (LAD), illustrating, for the first time,
such S-shaped diffusion patterns for local areas. This system level
analysis also reveals insights about the volatility of the adoption of
the Web. Then, exploratory spatial data analysis offers some first
indications of the importance of the different spatial diffusion
mechanisms in shaping the diffusion of the Web in the UK. Importantly,
the analysis adopts two distinct geographical scales: the LAD and the
most detailed scale of UK census areas, the Output Areas (OA). Lastly, a
novel methodological framework based on Machine Learning (ML) algorithms
is developed to test the role of such mechanisms at these two scales.
The results reveal, for the first time, the importance of different
spatial diffusion mechanisms in the diffusion of an intangible, digital
technology at very granular geographical scales.

The structure of the paper goes as follows. Section~\ref{sec-litreview}
reviews the key concepts in the diffusion literature as well as relevant
empirical studies. Then, Section~\ref{sec-datamethods} describes the
data and methods used in this paper. The results of the different stages
of the analysis are presented in Section~\ref{sec-results}. The
discussion of the results and the conclusions of the paper are offered
in Section~\ref{sec-conclusions}.

@@

\citet{hagerstrand1968innovation} also emphasised the role of the
top-down hierarchy of urban systems and highlighted that new
technologies and innovations are firstly adopted in larger cities and
then diffuse to second tier ones. This is a sequential instead of a
simultaneous process, which resembles the `lead-lag' spatial
acceleration effect in market research
\citep{bento2018time, PERES201091}. Schmidt's Law empirically
illustrates a similar pattern \citep{grubler1990rise}. Core and highly
agglomerated regions is where new technologies are invented and
commercially deployed. This is where the first adopters tend to be
based. Then, technologies spread to the rim and eventually to the
periphery. Although adoption pace might be higher when new technologies
finally arrive to the periphery, the saturation levels there may never
reach the ones in the core because of the lack of infrastructure or
other necessary institutions \citep{leibowicz2016representing}.

@@

Although the above processes have been studied in innovation studies,
the distancing of economic geographers from research questions about the
diffusion of new technologies and innovations resulted to a knowledge
gap regarding the explanatory power of these mechanisms at granular,
local scales \citep{ding2010modeling}. As discussed in
Section~\ref{sec-introduction}, such spatially explicit knowledge can
directly inform local technology and infrastructure policies. Indeed,
while the spatial dimension is present in the following studies, the
geographical level of analysis is often too coarse to ascertain the role
of the above mechanisms at the local scale. \citet{beardsell1999spatial}
studied the evolution of the computer industry in 317 US metropolitan
areas during the 1977-1992 period using employment data. Their analysis
indicated that the relative size distribution holds for urban computer
employment and also urban heterogeneity is essential in explaining this
distribution. In a recent study, \citet{bednarz2020pulled} focused on
wind turbines and modelled their spatial diffusion across 402 German
regions during 1970-2015. Their key finding is that local demand than
local supply was the main driving factor. \citet{haller2011determinants}
employed firm panel data for Ireland during 2003-2005 and illustrated
that when firms are located in a region or industry with high share of
firms having a website, they are more likely to have a website too. At a
global scale \citet{perkins2011internet} explored whether the adoption
of previous communication technologies that is mail, telegrams and
telephones were shaped by similar socioeconomic factors as the Internet.
Their results indicated common patterns regarding the drivers behind the
adoption of different communication technologies.
\citet{feldman2015rknowledge} focused on inventions and used patent data
to map how the the rDNA technology diffused among inventors across 366
US metropolitan areas. \citet{ding2010modeling} modelled the spatial
diffusion of mobile telecommunications across the highest level of
administrative division in China (29 regions in total). Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
\citet{bakher2013diffusion} compared different the growth curves as a
means to model the diffusion of online shopping in Australia during the
1998-2009 period, but they did not consider any spatial mechanisms. At a
global scale \citet{PAPAGIANNIDIS2015308} modelled the diffusion of
different web technologies and practices across countries. Using similar
but less extensive data as the one used here, their analysis illustrated
how the diffusion of different web technologies and practices follows an
S-shaped pattern as well as the different diffusion rates of the
different technologies and practices. The most detailed consideration of
the spatial footprint of the diffusion of a new technology comes from
\citet{lengyel2020role}, who analysed the adoption and churn of a
Hungarian online social network. Their results were in agreement with
the early theoretical and empirical contributions reviewed here:
assortativity, urban scaling and distance are the key drivers of spatial
diffusion. Based on the above, the main empirical hypothesis states:

\begin{quote}
H\textsubscript{1}: The three diffusion mechanisms namely distance,
urban hierarchy and the the S-shaped pattern of the cummulative level of
adoption shaped the diffusion of the commercial Web in the UK even at
granular geographical scales.
\end{quote}

@@

\begin{quote}
H\textsubscript{2}: The diffusion of the commecrial Web is shaped by
both early adoption and late comer effects.
\end{quote}

These are the two hypotheses are empirically tested in
Section~\ref{sec-results} using the data and methods described in the
next section.

@@

To capture the diffusion of the commercial Web, a website density metric
is developed for two different geographical scales that are discussed
later in this section. The main data are counts of commercial websites
that are calculated using data from the Internet Archive\footnote{\href{See\%20https://archive.org/}{https://archive.org/}.}
and, specifically, the JISC UK Web Domain Dataset \citep{ukwebarchive}.
The Internet Archive is one of the most complete and oldest archive of
webpages in the world operating since 1996
\citep{ainsworth2011much, holzmann2016dawn}. It is a web crawler, which
discovers webpages by following the hyperlinks of every webpage it
archives. This dataset, which is curated by the British Library,
contains all the archived webpages from the UK country code top level
domain (ccTLD - .uk) from the 1996--2012 period. In essence, this is a
long list of 2.5 billion URLs of archived webpages including also the
archival timestamp.

@@

\noindent What is challenging is that this aggregation approach, which
has been used elsewhere \citep{tranosuk, shoreditch} may lead to
websites with multiple postcodes. As per the above example,
\href{http://www.website.co.uk}{www.website.co.uk} includes two unique
postcodes: B15 2TT and BS8 1SS. The distribution of the number of
postcodes per website for 2000 is presented in Table \ref{f2000}, which
illustrates a wide range. At the left end of the distribution, there are
websites anchored to a unique location (72\% of all the reconstructed
websites in 2000), which may represent a small company with a single
trading location. At the right end, there are websites with thousands of
different postcodes. Considering the time period of the analysis, such
cases can represent directories which used to be popular in the
pre-search engines early times of the commercial Internet as well as
real estate websites \citep{tranosuk}.

The analysis is firstly conducted on websites, which only contain one
unique postcode. As a robustness check, the analysis is replicated for
an extended subset of websites, which include up to 10 unique postcodes
to capture commercial websites which point to multiple locations. They
are geolocated by equally attaching them to these locations. This
extended sample includes 94 percent of all the archived websites in
2000.

@@

\caption{Number of unique postcodes per .co.uk website, 2000.\label{f2000}}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Postcodes & F & F (\%) & Cummulative F\\
\midrule
(0,1] & 41,596 & 0.718 & 0.718\\
(1,2] & 6,451 & 0.111 & 0.830\\
(2,10] & 6,163 & 0.106 & 0.936\\
(10,100] & 2,975 & 0.051 & 0.988\\
(100,1000] & 646 & 0.011 & 0.999\\
\addlinespace
(1000,10000] & 62 & 0.001 & 1.000\\
(10000,100000] & 4 & 0.000 & 1.000\\
\bottomrule
\multicolumn{4}{l}{\rule{0pt}{1em}\textit{Source: } Tranos et al. 2021}\\
\end{tabular}
\end{table}

Another data cleaning step deals with some extreme outliers. Figure
\ref{correct} plots the website counts for the top 1000 postcodes. Some
obvious outliers can be observed for the 2002-2006 period for a handful
of postcodes, which can be attributed to link farms \citep{BL2013links}.
The website counts for these postcodes (SE24 9HP, CV8 2ED, GL16 7YA, CW1
6GL, M28 2SL, DE21 7BF), which in 2004 or 2005 had more than 1000
websites pointing to them, were replaced with predicted values based on
a simple panel regression model with postcode fixed effects and yearly
dummy variables. These postcodes refer to a residential area with a
small park in South London, residential areas outside Warwick and
Manchester, a rural area in the south border of England and Wales and
small business parks in Crew and Derby. Hence, it is difficult to
believe that these postcodes genuinely hosted such a large amounts of
websites, which were later extensively declined. To put the magnitude of
the data imputation into perspective, this process affected 6 out of the
557,808 postcodes included in the data. As the Section~\ref{sec-results}
illustrates, this led to a small increase of the model predictive power.

@@

{\centering \includegraphics[width=1\textwidth,height=0.7\textheight]{tranos2025_files/figure-pdf/unnamed-chunk-2-1.pdf}

@@

the Local Authority Districts (LAD) and the Output Areas (OA). The
former is an administrative unit and there are 374 LAD in the UK. The
latter is the smallest census-based geographical unit and there are
c.~230,000 of them in the UK. This methodological choice will allow the
mapping of the diffusion of web technologies and the assessment of the
diffusion mechanisms at these two very different spatial scales.

To create the website density metric, the yearly website counts at the
LAD level are standardised by the number of firms in LAD to avoid biases
associated with LAD hosting a large number of firms. Given that there is
no such readily available statistic for the OA, the actual OA level
counts are used. However, because of the the consistent spatial
definition of OA (they host 40-250 households),\footnote{\url{https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies}}
website counts in OA are interpreted as a density metric too.

These website density metrics are used for the three different stages of
the analysis. Firstly, a system-level analysis explores how the
diffusion of web technologies in the UK fits with the well-established
S-curve of cumulative adoption. To do so, the following logistic
function (Equation \ref{eq:s}) is estimated for the whole of the UK and
for each LAD separately:

@@

\noindent \(k\) is the asymptote or, in other words, the saturation
level, \(b\) the overall growth rate, and \(t_{0}\) the \emph{inflection
point} of maximum growth at \(k/2\), where the logistic function is
symmetrical \citep{wilson201281}. The \(t_0\) of each LAD is compared
against the \(t_0\) of the UK to delineate whether a LAD reached that
point faster or slower than the country average. Importantly, an
accuracy criterion was imposed and only LAD with \(R^2 > 0.9\) were
included in this analysis. To estimate Equation \ref{eq:s} a
self-starting logistic growth model was employed using the \texttt{nls}
and \texttt{SSlogis} functions in \texttt{R}. This logistic growth model
depicts the initial period of slow diffusion of a new technology, which
is followed by the fast, exponential growth stage to then eventually
slow down and saturate \citep{wilson201281, grubler1999dynamics}.

The system-level analysis also focuses on the volatility of the web
adoption to depict places with high concentration of early adopters and,
equally, latecomers. To do so, the change of the ranking of the UK LAD
over time is plotted and discussed. Both the S-curve and the volatility
analysis focus only on LAD as the very large number of OA would have
made such analysis difficult to visualise and interpret.

Next, exploratory spatial data analysis depicts whether the two main
drivers of spatial diffusion -- namely neighbourhood and hierarchy --
underpin the diffusion of the Web in the UK. To capture the former and
following \citet{ding2010modeling} the Moran's I and the Local Index of
Spatial Association (LISA) are estimated for the website density. To
address the hierarchy mechanism, the Gini coefficient -- an established
metric of inequality -- is calculated. All the above are computed and
plotted longitudinally both for the LAD and the OA.

Lastly, a modelling framework is developed to test the above diffusion
mechanisms. The overarching aim is to build a model that can test how
well these mechanisms can predict website density:

\begin{align}
Website\,Density_{t} \sim Hierarchy_{t-1} + Neighbourhood_{t-1} + S-curve_{t}\label{eq:rf.generic}
\end{align}

\noindent Equation \ref{eq:rf.generic} is estimated using Random Forest
(RF). This is a popular ML algorithm for both regression and
classification problems \citep{biau2012analysis}. It was introduced by
\citet{breiman2001random} and has become a go-to data science tool. RF
can effectively handle skewed distributions and outliers, model
non-linear relationships, require minimal hyperparameter tuning, exhibit
low sensitivity to these parameters, and have relatively short training
times \citep{Caruana2008, liaw2002classification, yan2020using}. These
attributes match well with the website density data characteristics
including skewness especially for the OA. Also, the large data size (c.
230k data points for each of the 17 years) calls for fast training
times. Importantly, RF predictions tend to be more accurate than those
from single regression trees and outperform Ordinary Least Squares in
out-of-sample predictions, even with moderate-sized training data and a
small number of predictors
\citep{mullainathan2017machine, athey2019machine, sulaiman2011intelligent, pourebrahim2019trip, biau2012analysis}.

>>>>

\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A multi-scale story of the diffusion of a new technology: the web},
  pdfkeywords={diffusion, technologies, Web, digital},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

@@

\begin{abstract}
This paper investigates the spatial diffusion of a new technology that
is the Web in the UK. It employs novel data and machine learning methods
to model the influence of well-established diffusion mechanisms.
Contrary to previous studies, it adopts multiple scales, high spatial
granularity and a long study period that captures the early stages of
the Web until its maturity (1996-2012). Findings reveal the importance
of such spatial mechanisms (namely distance, urban hierarchy and the
S-shaped pattern of the cumulative level of adoption) even at granular
scales. They also highlight spatial heterogeneity and instances of
leapfrogging.
\end{abstract}

@@

Economic geographers were always interested in how new technologies and
innovations diffuse across space and time with the work of
\citet{hagerstrand1968innovation}, which demonstrated the importance of
communications as a mechanism for the diffusion of innovations, being
the most impactful. It was followed by various empirical studies -- see
for instance \citet{ormrod1990} and \citet{iso2005}. However, the torch
of exploring and modelling such diffusion processes had been passed to
other disciplines such as economics, business studies and sociology well
before the `cultural turn' of economic geography
\citep{perkins2005international, ding2010modeling}. As a consequence,
relevant studies did not investigate diffusion mechanisms at granular
geographical scales to provide local policy insights, for instance,
regarding infrastructural development and local technology policies. A
potential explanation of the lack of geographical studies exploring the
diffusion of new and, more specifically for this paper, digital
technologies across both space and time can be attributed to the
scarcity of relevant and granular enough data, a problem also
highlighted by \citet{iso2005} and \citet{kemeny2011international}. As
\citet{zook2022mapping} pointed, digital activities are hardly ever
captured in official data.

This paper offers such a contribution: an economic geography study
revealing the importance of spatial mechanisms for the diffusion of a
digital technology that is the Web in the UK at a high level of spatial
granularity from the very early days until its maturity (1996-2012). It
specifically focuses on the diffusion of commercial websites. It does so
by employing a novel source of big data and machine learning algorithms
to model the \emph{active} engagement with Web -- that is creating and
maintaining instead of just browsing a website -- during that period and
tests how different spatial and temporal mechanisms shaped its
diffusion. This paper also exemplifies how the combination of data
sources and state-of-the-art research methods which escape the
traditional social science domain can offer new lenses to geographical
research regarding the understanding of technological diffusion.

The motivation for this paper lies in the fact that there are various
stakeholders who are interested in knowing how new digital technologies
diffuse over space and time and use this knowledge to make predictions
about the diffusion of related future technologies. As per
\citet{leibowicz2016representing}, historical studies agree that
technologies diffuse differently in terms of times, rates, and
geographies and can be driven by related policies \citep{victor1993}.
\citet{meade2021modelling} highlight that a variety of actors have a
direct interest in gaining such knowledge including network equipment
suppliers, network operators, regulatory and local authorities. These
processes and their effects vary a lot across scales: although the
diffusion of a new technology might not be optimal at a local level, it
might be beneficial from a global perspective as it could lead to faster
diffusion to less advantaged places \citep{leibowicz2016representing}.
Despite the spatial heterogeneity of such diffusion mechanisms and the
policy relevance, there is a scarcity of studies analysing the diffusion
of new technologies at a detailed geographical level because of the lack
of detailed enough data and adequate research methods.

@@

Another line of motivation has to do with the very nature of the Web as
a digital technology and how it evolved over time. The digital
geographies field has debunked premature claims about the Internet's
placeless-ness by mapping various of its facets and technological layers
\citep{tranos2013death}, but the lack of relevant long time-series data
prevented researchers from observing digital phenomena overtime. Most of
these studies focused on its hard infrastructure
\citep[e.g.][]{malecki2002economic, moss2000internet}, on Internet
subscribers \citep[e.g.][]{blank2018local}, or on social media users
\citep[e.g.][]{crampton2013beyond}. Exceptions include early work from
\citet{zook2000web}, \citet{zook2001old} and \citet{moss1997tracking},
which utilised domain name registration data and use it to map and
illustrate the spatial footprint of the web content. Pivotal as these
contributions have been in understanding the economic geographies of the
Internet, they were based on cross-sectional, often spatially detailed,
mapping and on short temporal comparisons (e.g.~1993-1998) and did not
observe the evolution and diffusion of the Web until its maturity. As we
know from the relevant literature, different technologies diffuse at
different rates and there is no theory that can explain these
differences \citep{leibowicz2016representing}. Hence, mapping and
modelling the diffusion of the Web can offer useful insights.

@@

To analyse these data, a diverse set of methods are employed. Firstly,
cumulative adoption curves are estimated for the UK as a whole and for
the Local Authority Districts (LAD), illustrating, for the first time,
such S-shaped diffusion patterns for local areas. The exploratory
spatial data analysis offers some first indications of the importance of
the different spatial diffusion mechanisms in shaping the diffusion of
the Web in the UK. Importantly, the analysis adopts two distinct
geographical scales: the LAD and the most detailed scale of UK census
areas, the Output Areas (OA). Then, a novel methodological framework
based on Machine Learning (ML) algorithms is developed to test the role
of such mechanisms at these two scales. The results reveal, for the
first time, the importance of different spatial diffusion mechanisms in
the diffusion of a digital technology at very granular geographical
scales. The analysis also reveals that the diffusion of the Web is
characterised by spatial heterogeneity, well-established spatial
patterns and both stability and volatility.

The structure of the paper goes as follows. Section~\ref{sec-litreview}
reviews the mechanisms that shape the diffusion of a new technology and
sets up the two hypotheses that are empirically tested in
Section~\ref{sec-results}. Section~\ref{sec-datamethods} describes the
data and methods used in this paper. The discussion of the results and
the conclusions of the paper are offered in
Section~\ref{sec-conclusions}.

@@

\citet{hagerstrand1968innovation} also emphasised the role of the
top-down hierarchy of urban systems and highlighted that new
technologies and innovations are firstly adopted in larger cities and
then diffuse to second tier ones. This is a sequential instead of a
simultaneous process, which resembles the `lead-lag' spatial
acceleration effect in market research
\citep{bento2018time, PERES201091}. Schmidt's Law empirically
illustrates a similar pattern \citep{grubler1990rise}. Core and highly
agglomerated regions is where new technologies are invented and
commercially deployed. This is where the first adopters tend to be
based. Then, technologies spread to the rim and eventually to the
periphery. Although adoption pace might be higher when new technologies
finally arrive to the periphery, the saturation levels may never reach
the ones in the core because of the lack of infrastructure or other
necessary institutions \citep{leibowicz2016representing}.

@@

Although the above processes have been studied in innovation studies,
the distancing of economic geographers from research questions about the
diffusion of new technologies and innovations resulted to a knowledge
gap regarding the explanatory power of these mechanisms at granular,
local scales \citep{ding2010modeling}. As discussed in
Section~\ref{sec-introduction}, such spatially explicit knowledge can
directly inform local technology and infrastructure policies. Indeed,
while the spatial dimension is present in the following studies, the
geographical level of analysis is often too coarse to ascertain the role
of the above mechanisms at the local scale. \citet{beardsell1999spatial}
studied the evolution of the computer industry in 317 US metropolitan
areas during the 1977-1992 period using employment data. Their analysis
indicated that the relative size distribution holds for urban computer
employment and also urban heterogeneity is essential in explaining this
distribution. In a recent study, \citet{bednarz2020pulled} focused on
wind turbines and modelled their spatial diffusion across 402 German
regions during 1970-2015. Their key finding is that local demand than
local supply was the main driving factor. \citet{haller2011determinants}
employed firm panel data for Ireland during 2003-2005 and illustrated
that when firms are located in a region or industry with high share of
firms having a website, they are more likely to have a website too. At a
global scale \citet{perkins2011internet} explored whether the adoption
of previous communication technologies that is mail, telegrams and
telephones were shaped by similar socioeconomic factors as the Internet.
Their results indicated common patterns regarding the drivers behind the
adoption of different communication technologies.
\citet{feldman2015rknowledge} focused on inventions and used patent data
to map how the the rDNA technology diffused among inventors across 366
US metropolitan areas. \citet{ding2010modeling} modelled the spatial
diffusion of mobile telecommunications across the highest level of
administrative division in China (29 regions in total). Their analysis
indicated that socioeconomic characteristics are important determinants
of the timing, speed and the level of mobile diffusion within China.
\citet{bakher2013diffusion} compared different growth curves as a means
to model the diffusion of online shopping in Australia during the
1998-2009 period, but they did not consider any spatial mechanisms. At a
global scale \citet{PAPAGIANNIDIS2015308} modelled the diffusion of
different web technologies and practices across countries. Using similar
but less extensive data as the one used here, their analysis illustrated
how the diffusion of different web technologies and practices follows an
S-shaped pattern as well as the different diffusion rates of the
different technologies and practices. The most detailed consideration of
the spatial footprint of the diffusion of a new technology comes from
\citet{lengyel2020role}, who analysed the adoption and churn of a
Hungarian online social network. Their results were in agreement with
the early theoretical and empirical contributions reviewed here:
assortativity, urban scaling and distance are the key drivers of spatial
diffusion. Based on the above, the main empirical hypothesis that is
tested in this paper states:

\begin{quote}
H\textsubscript{1}: The three diffusion mechanisms namely distance,
urban hierarchy and the S-shaped pattern of the cumulative level of
adoption shaped the diffusion of the commercial Web in the UK even at
granular geographical scales.
\end{quote}

@@

\begin{quote}
H\textsubscript{2}: The diffusion of the commercial Web is shaped by
both early adoption and latecomer effects.
\end{quote}

These two hypotheses are empirically tested in Section~\ref{sec-results}
employing the data and methods described in the next section.

@@

To capture the diffusion of the commercial Web, a website density metric
is developed for two different geographical scales that are discussed
later in this section. The main data are counts of commercial websites
that are calculated using data from the Internet Archive\footnote{\url{https://archive.org/}.}
and, specifically, the JISC UK Web Domain Dataset \citep{ukwebarchive}.
The Internet Archive is one of the most complete and oldest archive of
webpages in the world operating since 1996
\citep{ainsworth2011much, holzmann2016dawn}. It is a web crawler, which
discovers webpages by following the hyperlinks of every webpage it
archives. This dataset, which is curated by the British Library,
contains all the archived webpages from the UK country code top level
domain (ccTLD - .uk) from the 1996--2012 period. In essence, this is a
long list of 2.5 billion URLs of archived webpages including also the
archival timestamp.

@@

\noindent What is challenging is that this aggregation approach, which
has been used elsewhere \citep{tranosuk, shoreditch} may lead to
websites with multiple postcodes. As per the above example,
\href{http://www.website.co.uk}{www.website.co.uk} includes two unique
postcodes: B15 2TT and BS8 1SS. The distribution of the number of
postcodes per website for 2000 is presented in Table \ref{f2000}, which
illustrates a wide range. At the left end of the distribution, there are
websites anchored to a unique location (72\% of all the reconstructed
websites in 2000), which may represent a small company with a single
trading location. At the right end, there are websites with thousands of
different postcodes. Considering the time period of the analysis, such
cases can represent directories which used to be popular in the
pre-search engines early times of the commercial Internet as well as
real estate websites \citep{tranosuk}. Previous research which used the
same methodology for a small subset of these data illustrates the
robustness of the above approach: its results matched in terms of
accuracy past rounds of research based on extensive qualitative work,
interviews and observational studies for the same study area
\citep{shoreditch}.

The analysis is firstly conducted on websites, which only contain one
unique postcode. As a robustness check, the analysis is replicated for
an extended subset of websites, which include up to 10 unique postcodes
to capture commercial websites which point to multiple locations. They
are geolocated by equally attaching them to these locations. This
extended sample includes 94 percent of all the archived websites in 2000
(see Table \ref{f2000}).

@@

\caption{Number of unique postcodes per .co.uk website, 2000.\label{f2000}}
\centering
\begin{tabular}[t]{lrrr}
\toprule
Postcodes & F & F (\%) & cumulative F\\
\midrule
(0,1] & 41,596 & 0.718 & 0.718\\
(1,2] & 6,451 & 0.111 & 0.830\\
(2,10] & 6,163 & 0.106 & 0.936\\
(10,100] & 2,975 & 0.051 & 0.988\\
(100,1000] & 646 & 0.011 & 0.999\\
\addlinespace
(1000,10000] & 62 & 0.001 & 1.000\\
(10000,100000] & 4 & 0.000 & 1.000\\
\bottomrule
\multicolumn{4}{l}{\rule{0pt}{1em}\textit{Source: } Tranos et al. 2021}\\
\end{tabular}
\end{table}

Another data cleaning step deals with some extreme outliers. Figure
\ref{correct} plots the website counts for the top 1000 postcodes. Some
obvious outliers can be observed for the 2002-2006 period for a handful
of postcodes, which can be attributed to link farms \citep{BL2013links}.
The website counts for these postcodes (SE24 9HP, CV8 2ED, GL16 7YA, CW1
6GL, M28 2SL, DE21 7BF), which in 2004 or 2005 had more than 1,000
websites pointing to them, were replaced with predicted values based on
a simple panel regression model with postcode fixed effects and yearly
dummy variables. These postcodes refer to a residential area with a
small park in South London, residential areas outside Warwick and
Manchester, a rural area in the south border of England and Wales and
small business parks in Crew and Derby. Hence, it is difficult to
believe that these postcodes genuinely hosted such a large amounts of
websites, which were later extensively declined. To put the magnitude of
the data imputation into perspective, this process affected 6 out of the
557,808 postcodes included in the data. As Section~\ref{sec-results}
illustrates, this led to a small increase of the model predictive power.

@@

{\centering \includegraphics[width=1\textwidth,height=0.7\textheight]{anonymised_files/figure-pdf/unnamed-chunk-2-1.pdf}

@@

To empirically test the two hypotheses stated in
Section~\ref{sec-litreview}, the website density metric was calculated
for two different scales the Local Authority Districts (LAD) and the
Output Areas (OA). The former is the lowest administrative unit in the
UK and there are 374 LAD. The latter is the smallest census-based
geographical unit and there are c.~230,000 of them in the UK. To create
the website density metric, the yearly website counts at the LAD level
are standardised by the number of firms in LAD to avoid biases
associated with LAD hosting a large number of firms. Given that there is
no such readily available statistic for the OA, the actual OA level
counts are used. However, because of the the consistent spatial
definition of OA (they host 40-250 households),\footnote{\url{https://www.ons.gov.uk/methodology/geography/ukgeographies/statisticalgeographies}}
website counts in OA are interpreted as a density metric too.

These website density metrics are used for the different stages of the
analysis. Starting with the first hypothesis (H\textsubscript{1}), to
test whether the S-curve accurately describes the cumulative adoption of
the Web at a local scale, the following logistic function (Equation
\ref{eq:s}) is estimated for the whole of the UK and then for each LAD
separately:

@@

\noindent \(k\) is the asymptote or, in other words, the saturation
level, \(b\) the overall growth rate, and \(t_{0}\) the \emph{inflection
point} of maximum growth at \(k/2\), where the logistic function is
symmetrical \citep{wilson201281}. The \(t_0\) of each LAD is compared
against the \(t_0\) of the UK to delineate whether a LAD reached that
point faster or slower than the country average. This heterogeneity is
discussed further when testing the second hypothesis
(H\textsubscript{2}). Importantly, an accuracy criterion was imposed and
only LAD with \(R^2 > 0.9\) were included in this analysis. To estimate
Equation \ref{eq:s} a self-starting logistic growth model was employed
using the \texttt{nls} and \texttt{SSlogis} functions in \texttt{R}.
This logistic growth model depicts the initial period of slow diffusion
of a new technology, which is followed by the fast, exponential growth
stage to then eventually slow down and saturate
\citep{wilson201281, grubler1999dynamics}.

Next, exploratory spatial data analysis provides some first insights of
whether the other two drivers of spatial diffusion -- namely
neighbourhood and hierarchy -- underpin the diffusion of the Web in the
UK. To capture the former and following \citet{ding2010modeling} the
Moran's I and the Local Index of Spatial Association (LISA) are
estimated for the website density. To capture the hierarchy mechanism,
the Gini coefficient -- an established metric of inequality -- is
calculated. All the above are computed and plotted longitudinally both
for the LAD and the OA.

To formally test H\textsubscript{1} a modelling framework is developed
to assess the above diffusion mechanisms. The overarching aim is to
build a model that can test how well these three mechanisms can predict
website density:

\begin{align}
Website\,Density_{t} \sim Hierarchy_{t-1} + Neighbourhood_{t-1} + S\text-curve_{t}\label{eq:rf.generic}
\end{align}

To capture the hierarchical mechanism the model include as predictors a
one year lag of website density in London, the largest city in the UK,
and a one year lag of the website density in the nearest city. Both
variables clearly illustrate the urban hierarchy in the UK. Especially
for the OA models, a third predictor is added to capture hierarchy at a
much more granular scale: a one year lag of the website density in the
the nearest retail centre \footnote{The retail centres data have been
  obtained by the Consumer Data Research Centre, an ESRC data
  investment, under project ID CDRC 498-01, ES/L011840/1; ES/L011891/1:
  \url{https://data.cdrc.ac.uk/dataset/retail-centre-boundaries-and-open-indicators}.}.
As the literature illustrates, retail centres are key components of
urban systems \citep{dennis2002central, jones2021reframing}. Due to the
small sizes of the retail centres, the latter is only relevant for and
included in the OA models and not in the LAD models. In addition, the
model includes the distance to London, the nearest city and the nearest
retail centre. The underlying logic is that the level of website
adoption in a spatial unit depends on the level of the adoption in
places further up in the urban hierarchy the previous year. The
inclusion of the distance variables incorporates spatial structure into
the hierarchy argument. To depict the neighbourhood mechanism, the
website density of the neighbouring spatial units in the previous year
is employed. This is calculated using rook continuity. Again, the
underpinning rationale is that the level of web adoption within a
spatial unit depends on the level of web adoption in the neighbouring
spatial units the year before. This represents the `hitting nearby
locations first' argument. Therefore, the spatial and temporal lag of
the website density in LAD and OA is calculated. Lastly, the temporal
pattern of the cumulative adoption is captured by a time trend variable.
Hence, the model will follow the following generic form (Eq.
\ref{eq:model}):
